---
title: "Data Simulation & Hierarchical Models: Day 2"
date: 2025-08-13
---

# Review and questions

- Plotting residuals and residuals vs. fitted
- Questions?
- Exercises?

# Outline for today

- New data: thanks, Carmen!
- Transformations
- Dealing with extreme/weird values
- Categorical predictors
- Multiple predictors
- Interactions
- Model comparisons vs. inference

# Load packages

```{r}
library(MASS)
library(tidyverse)
library(HistData)
```


# New data: child study from Carmen

  - **IMPORTANT**: Carmen has asked that we do not share this data outside the workshop.
  
```{r}
carmen <- read_csv("summer25_stat_workshop_Canphon_parti24.csv")
colnames(carmen)

```


# Exploring, Transforming, etc.

```{r}
ggplot(carmen, aes(TotalLook)) + geom_histogram()

ggplot(carmen, aes(log(TotalLook))) + geom_histogram()

carmen |> group_by(SLx, GapType, Age) |> summarize(count = n()) |> print(n = 50)

carmen <- filter(carmen, ! SLx %in% "sup")

carmen |> filter(TotalLook > 79950) |> group_by(SLx, GapType, Age) |> summarize(count = n()) |> print(n = 50)

carmen |> group_by(SLx, GapType, Age) |> 
  summarize(meanLooking = mean(TotalLook))

carmen$logLook <- log(carmen$TotalLook)
```

## Start looking at predictors


```{r}

```

