---
title: "Data Simulation & Hierarchical Models: Day 4"
date: 2025-08-14
---

# Review and questions

- Questions?
- Scott's homework: checking out the `contrast` package

# Outline for today

- Generate a random effect
- Use `lmer()` to fit the model
- Practice using `lmer()` on Carmen's data
- Interpreting random effects
- Comparing models with different random effects
- Practice: London's data

# Loading packages

```{r}
library(tidyverse)
library(lme4)
library(lmerTest)
library(HistData)
```

# Recapping simulation from a simple linear model

```{r}
galton_fit <- lm(child ~ 1 + parent, Galton)
summary(galton_fit)
(parent_mean <- mean(Galton$parent))
(parent_sd <- sd(Galton$parent))

intercept <- 24
slope <- 0.6
rmse <- 2.25

n_fake <- 1000
fake_data <- data.frame(parent = rnorm(n_fake, mean = 68, sd = 1.8))
fake_data$child <- intercept + slope * fake_data$parent + rnorm(n_fake, mean = 0, sd = rmse)

summary(lm(child ~ 1 + parent, fake_data))

generate_height_data <- function(n, intercept, slope, rmse,
                                 parent_mean, parent_sd) {
  fake_data <- data.frame(parent = rnorm(n, mean = parent_mean, sd = parent_sd))
  fake_data$child <- intercept + slope * fake_data$parent + rnorm(n, mean = 0, sd = rmse)
  
  return(fake_data)
}

fake_data2 <- generate_height_data(1000, 25, 0.6, 2.2, 68, 1.8)

summary(lm(child ~ 1 + parent, fake_data2))

```

Now let's do random effects

```{r}
n_countries <- 25

(country_parent_means <- rnorm(n = n_countries, mean = 68, sd = 2))
(country_intercepts <- rnorm(n = n_countries, mean = 24, sd = 3))

n_per_country <- 40
full_height_data <- data.frame(parent = NULL, child = NULL, country = NULL)
for(country in 1:n_countries) {
  
  country_data <- generate_height_data(n_per_country,
                                       intercept = country_intercepts[country],
                                       slope = slope,
                                       rmse = rmse,
                                       parent_mean = country_parent_means[country],
                                       parent_sd = 2)
  country_data$country <- country
  full_height_data <- bind_rows(full_height_data, country_data)
  
}

head(full_height_data)
full_height_data |> group_by(country) |> summarize(count = n())
cell_means <- full_height_data |> group_by(country) |> summarize(mean_parent = mean(parent)) 
cell_means$theoretical <- country_parent_means
head(cell_means)

```

